A weakly structured stem for human origins in Africa

### 2022-02-13 paper.tex
Update paper.tex

\title{A weakly structured stem for human origins in Africa}
\author[1]{Aaron P. Ragsdale}
\author[2]{Timothy D. Weaver}
\author[3]{Elizabeth G. Atkinson}
\author[4]{Eileen Hoal}
\author[4]{Marlo M\"{o}ller}
\author[2,6,*]{Brenna M. Henn}
\author[7,**]{Simon Gravel}
\affil[1]{Department of Integrative Biology, University of Wisconsin, Madison, WI, USA}
\affil[2]{Department of Anthropology, University of California, Davis, Davis, CA, USA}
\affil[3]{FIXME: }
\affil[4]{DSI-NRF Centre of Excellence for Biomedical Tuberculosis Research; South African Medical Research Council Centre for Tuberculosis Research; Division of Molecular Biology and Human Genetics, Faculty of Medicine and Health Sciences, Stellenbosch University, Cape Town, South Africa}
\affil[6]{UC Davis Genome Center, University of California, Davis, Davis, CA, USA}
\affil[7]{Department of Human Genetics, McGill University, Montreal, QC, Canada}
\affil[*]{bmhenn@ucdavis.edu}
\affil[**]{simon.gravel@mcgill.ca}
\maketitle

\abstract{
While a modern human origin within Africa is now broadly accepted, considerable
uncertainty surrounds specific models of divergence and migration across the
continent. Progress is hampered by a paucity of fossil and genomic data, as
well as variability in dating. Here we use linkage disequilibrium and
diversity-based statistics, optimized for rapid, complex demographic inference
to discriminate among such models. We infer detailed demographic models for
populations across Africa, including representatives from eastern and western
groups, as well as 44 newly whole-genome sequenced individuals from the Nama
(Khoe-San). Despite the complexity of African population history, present-day
population structure dates back to Marine Isotope Stage (MIS) 5. The earliest
population divergence among contemporary populations occurs 120-135kya, between
the Khoe-San and other groups. Prior to the divergence of contemporary African
groups, we infer long-lasting structure between two or more weakly
differentiated ancestral Homo spp. populations connected by gene flow over
hundreds of thousands of years (i.e. a weakly structured stem). We find that
weakly-structured-stem models provide more likely explanations of polymorphism
that had previously been attributed to contributions from archaic hominins in
Africa.
In contrast to models with archaic introgression, we predict that fossil
remains from coexisting ancestral populations should be morphologically
similar.
Despite genetic similarity between these populations, an inferred 1--4\%
of genetic differentiation among contemporary human populations can be
attributed to genetic drift between stem populations.
We show that model misspecification explains variation in previous
divergence time estimates and argue that studying a suite of models is key to
robust inferences about deep history.
}

@@

\begin{figure}[ht!]
    \centering
    \includegraphics{figures/fig1.pdf}
    \caption{
        \textbf{Geography and genetic diversity within Africa.}
        (A) The combined 1000 Genomes and AGRP datasets include populations from
        West, East, and South Africa. Both PCA (B, C) and \texttt{ADMIXTURE}
        (D) illustrate the well-characterized high levels of genetic diversity
        and signatures of recent migration and gene flow within Africa.
        We built parameterized models using the newly-sequenced Nama, Mende
        from Sierre Leone, Gumuz, Oromo and Amhara from Ethiopia, and the
        British and Vindija Neanderthal individual (bold fonts).
        \aprcomment{work on this caption}
    }
    \label{fig:1}
\end{figure}

@@

Regardless of the model choice for early epochs, inference of human demographic
history for the last 150kya was remarkably robust. The earliest divergence among
contemporary human populations differentiates the southern African Nama from
other African groups between 110--135kya, with low to moderate levels of
subsequent gene flow (Table\ref{tab:migration-rates}). In none of the
high-likelihood models which we explored did the divergence between Nama and
other populations exceed $\sim$140kya. We conclude that
geographic patterns of contemporary \emph{Homo sapiens} population structure
date back to the Late Middle Stone Age in Africa, likely arising during MIS 5.
%\sgcomment{Alt formulation: Thus all models suggest that broad geographic patterns of contemporary population structure date back to MIS 5}
Although we find evidence for earlier population structure in Africa (see
below), present-day populations cannot be easily mapped onto these more ancient
stem groups as only a small proportion of drift between modern populations can
be attributed to drift between stems (Figures~\ref{fig:3} and
\ref{fig:supp-f4s-single-origin}--\ref{fig:supp-f4s-merger-with-stem-migration}).

Given this consistency in inferred recent history and the numerical challenge
of optimizing a large number of parameters, we fixed several parameters related
to recent population history so as to focus on more ancient events (Supp.
Methods). The time of divergence of Western and Eastern African populations was
set to 60kya, just prior to the split of Eurasians and East Africans 50kya. All
models infer relatively high subsequent gene flow between the two regions
($m\approx2\times10^{-4}$). Admixture from Neanderthals to Europeans directly
following the out-of-Africa migration was set to 1.5\% at 45kya (Supp.
Methods). Back-to-Africa gene flow at the beginning of the Holocene primarily
\sgcomment{Here it is not clear whether we are still talking about parameters that we fixed. I think we are also missing a clear description of the best-fit model. That *could* go after the description of the model uncertainty}
affected the ancestors of the Ethiopian agricultural populations, comprising
over half of their genetic ancestry, estimated to be 64--65\%. The past 5,000
years also saw major demographic changes, including strong population growth
for Western Africans as they specialized in yam and oil palm agriculture
(estimated 3-fold growth). We observe significant gene flow from the Amhara and
Oromo into the Nama, a signal which is likely a proxy for the movement of
Eastern African caprid and cattle pastoralists
\citep{Henn2008-xo,Breton2014-xb}, here estimated to constitute a 25\% ancestry
contribution 2,000 ya. Colonial period admixture from Europeans into the Nama
was estimated at 15\%, similar to proportions inferred by \texttt{ADMIXTURE}
\citep{Alexander2009-sw} (Figure~\ref{fig:1}).

>>>>

\title{A weakly structured stem for human origins in Africa}
\author[1]{Aaron P. Ragsdale}
\author[2]{Timothy D. Weaver}
\author[3]{Elizabeth G. Atkinson}
\author[4]{Eileen Hoal}
\author[4]{Marlo M\"{o}ller}
\author[2,6,^,*]{Brenna M. Henn}
\author[7,^,**]{Simon Gravel}
\affil[1]{Department of Integrative Biology, University of Wisconsin, Madison, WI, USA}
\affil[2]{Department of Anthropology, University of California, Davis, Davis, CA, USA}
\affil[3]{Department of Molecular and Human Genetics, Baylor College of Medicine, TX, USA}
\affil[4]{DSI-NRF Centre of Excellence for Biomedical Tuberculosis Research; South African Medical Research Council Centre for Tuberculosis Research; Division of Molecular Biology and Human Genetics, Faculty of Medicine and Health Sciences, Stellenbosch University, Cape Town, South Africa}
\affil[6]{UC Davis Genome Center, University of California, Davis, Davis, CA, USA}
\affil[7]{Department of Human Genetics, McGill University, Montreal, QC, Canada}
\affil[^]{Co-Corresponding Authors}
\affil[*]{bmhenn@ucdavis.edu}
\affil[**]{simon.gravel@mcgill.ca}
\maketitle

\abstract{
While a modern human origin within Africa is now broadly accepted, considerable
uncertainty surrounds specific models of divergence and migration across the
continent. Progress is hampered by a paucity of fossil and genomic data, as
well as variability in dating. Here we use linkage disequilibrium and
diversity-based statistics, optimized for rapid, complex demographic inference
to discriminate among such models. We infer detailed demographic models for
populations across Africa, including representatives from eastern and western
groups, as well as 44 newly whole-genome sequenced individuals from the Nama
(Khoe-San). Despite the complexity of African population history, present-day
population structure dates back to Marine Isotope Stage (MIS) 5. The earliest
population divergence among contemporary populations occurs 120-135kya, between
the Khoe-San and other groups. Prior to the divergence of contemporary African
groups, we infer long-lasting structure between two or more weakly
differentiated ancestral \emph{Homo spp.} populations connected by gene flow over
hundreds of thousands of years (i.e. a weakly structured stem). We find that
weakly structured stem models provide more likely explanations of polymorphism
that had previously been attributed to contributions from archaic hominins in
Africa.
In contrast to models with archaic introgression, we predict that fossil
remains from coexisting ancestral populations should be morphologically
similar.
Despite genetic similarity between these populations, an inferred 1--4\%
of genetic differentiation among contemporary human populations can be
attributed to genetic drift between stem populations.
We show that model misspecification explains variation in previous
divergence time estimates and argue that studying a suite of models is key to
robust inferences about deep history.
}

@@

\begin{figure}[ht!]
    \centering
    \includegraphics{figures/fig1.pdf}
    \caption{
        \textbf{Geography and genetic diversity within Africa.}
        (A) The combined 1000 Genomes and AGRP datasets include populations from
        West, East, and South Africa. Both PCA (B, C) and \texttt{ADMIXTURE}
        (D) illustrate the well-characterized high levels of genetic diversity
        and signatures of recent migration and gene flow within Africa.
        We built parameterized models using the newly-sequenced Nama from South Africa, Mende
        from Sierre Leone, Gumuz, Oromo and Amhara from Ethiopia, and the
        British and Vindija Neanderthal individual (bold fonts).
        \aprcomment{work on this caption}
    }
    \label{fig:1}
\end{figure}

@@

Regardless of the model choice for early epochs, inference of human demographic
history for the last 150kya was remarkably robust. The earliest divergence among
contemporary human populations differentiates the southern African Nama from
other African groups between 110--135kya, with low to moderate levels of
subsequent gene flow (Table \ref{tab:migration-rates}). In none of the
high-likelihood models which we explored did the divergence between Nama and
other populations exceed $\sim$140kya. 
We conclude that geographic patterns of contemporary \emph{Homo sapiens} population structure
date back to the late Middle Stone Age in Africa, likely arising during MIS 5.
Although we find evidence for earlier population structure in Africa (see
below), present-day populations cannot be easily mapped onto these more ancient
stem groups as only a small proportion of drift between modern populations can
be attributed to drift between stems (Figures~\ref{fig:3} and
\ref{fig:supp-f4s-single-origin}--\ref{fig:supp-f4s-merger-with-stem-migration}).

Given this consistency in inferred recent history and the numerical challenge
of optimizing a large number of parameters, we fixed several parameters related
to recent population history so as to focus on more ancient events (see below). Fixed parameters included
divergence time the time of divergence between Western and Eastern African populations,
set to 60kya, just prior to the split of Eurasians and East Africans set to 50kya. 
We also fixed the amount of admixture from Neanderthals to Europeans directly
following the out-of-Africa migration which was set to 1.5\% at 45kya (Supp. Methods). 
All models infer relatively high subsequent gene flow between the two regions
($m\approx2\times10^{-4}$). Back-to-Africa gene flow at the beginning of the Holocene primarily
affected the ancestors of the Ethiopian agricultural populations, comprising
over half of their genetic ancestry, estimated to be 64--65\%. The past 5,000
years also saw major demographic changes, including strong population growth
for Western Africans as they specialized in yam and oil palm agriculture
(estimated 3-fold growth). We observe significant gene flow from the Amhara and
Oromo into the Nama, a signal which is likely a proxy for the movement of
Eastern African caprid and cattle pastoralists
\citep{Henn2008-xo,Breton2014-xb}, here estimated to constitute a 25\% ancestry
contribution 2,000 ya. Colonial period admixture from Europeans into the Nama
was estimated at 15\%, similar to proportions inferred by \texttt{ADMIXTURE}
\citep{Alexander2009-sw} (Figure~\ref{fig:1}).
